A fine granularity load balancing technique for MMOG servers using a kd-tree to partition the space

### 2009-09-11 textos/sbgames2009/kdtree.tex

%% Author and Affiliation (multiple authors). Use: and between authors
\author{Manuscript number 60381}
%\author{Carlos Eduardo B. Bezerra, João L. D. Comba, Cláudio F. R. Geyer\\
%				Instituto de Informática\\
%				Universidade Federal do Rio Grande do Sul\\
%				Av. Bento Gonçalves, 9500, Porto Alegre
%}
\contactinfo{60381@jems}
%\contactinfo{\{carlos.bezera, comba, geyer\}@inf.ufrgs.br
%}
%% Keywords that describe your work.
\keywords{MMOGs, load balancing, distributed server, kd-trees.}

@@

In order to reduce this cost, several decentralized solutions have been proposed. Some of them use peer-to-peer networks, such as \cite{schiele2007rpp,rieche2007ppb,hampel2006ppa,elrhalibi2005abm,iimura2004zfg,knutsson2004pps}. Others propose the use of a distributed server composed of low-cost nodes connected through the Internet, as in \cite{ng2002msa,chertov:olb,lee2003sdl,assiotis2006dam}. Anyway, in all these approaches, the ``world'', or virtual environment of the game is divided into regions and for each region is assigned a server -- or a group of peers to manage it, when using peer-to-peer networks. Each of these regions must have a content such that the load imposed on the corresponding server is not greater than its capacity.

@@

\cite{ahmed2008mol}, for example, propose a cell-oriented load balancing model. To balance the load, their algorithm finds, first, all clusters of cells that are managed by the overloaded server. The smallest cluster is selected and, from this cluster, it is chosen the cell which has the least interaction with other cells of the same server -- the interaction between two cells A and B is defined by the authors as the number of pairs of avatars interacting with each other, one of them in A and the other one in B. The selected cell is then transferred to the least loaded server, considering ``load'' as the bandwidth used to send send state updates to the players whose avatars are positioned in the cells managed by that server. This process is repeated until the server is no longer overloaded or there is no more servers capable of absorbing more load -- in this case, one option could be to reduce the frequency at which state update messages are sent to the players, as suggested by \cite{bezerra2008a3}.

@@

In \cite{bezerra2009lbs}, it is also proposed the division into cells. To perform the division, the environment is represented by a graph (Figure \ref{fig:graph}), where each vertex represents a cell. Each edge in the graph connects two vertices representing neighboring cells. The weight of a vertex is the server's bandwidth occupied to send state updates to the players whose avatars are in the cell represented by that vertex. The interaction between any two cells define the weight of the edge connecting the corresponding vertices. To form the regions, the graph is partitioned using a greedy algorithm: starting from the heaviest vertex, at each step it is added the vertex connected by the heaviest edge to any of the vertices already selected, until the total weight of the partition of the graph -- defined as the sum of the vertices' weights -- reaches a certain threshold related to the total capacity of the server that will receive the region represented by that partition of the graph.

@@

This kind of data structure is generally used in computer graphics. However, as in MMOGs there is geometric information -- such as the position of each avatar in the environment --, space partitioning trees can be used. Moreover, we cand find in the literature techniques for keeping the partitions defined by the tree with a similar ``load''. In \cite{luque2005bpc}, for example, it is sought to reduce the time needed to calculate the collisions between pairs of objects moving through space. The authors propose the use of a BSP (binary space partitioning) tree to distribute the objects in the scene (Figure \ref{fig:bsp}). Obviously, if each object of one pair is completely inserted in a different partition, they do not collide and there is no need to perform a more complex test for this pair. Assuming an initial division, it is proposed by the authors a dynamic readjustment of the tree as objects move, balancing their distribution on the leaf-nodes of the tree and, therefore, minimizing the time required to perform the collision detection. Some of the ideas proposed by the authors may be used in the context of load balancing between servers in MMOGs.

@@

The load balancing approach proposed here is based on two criteria: first, the system should be considered heterogeneous (i.e. each server has a different amount of resources) and, second, the load on each server is \emph{not} proportional to the number of players connected to it, but to the amount of bandwidth required to send state update messages to them.

This choice is due to the fact that each player sends commands to the server at a constant rate, so the number of messages received by the server per unit time grows linearly to the number of players, whereas the number of state update messages sent by the server may be quadratic, in the worst case.

@@

Each node of the tree represents a region of the space and, moreover, in this node it is stored a split coordinate. Each one of the two children of that node represents a subdivision of the region represented by the parent node, and one of them represents the sub-region before the split coordinate and the other one, the sub-region containing points whose coordinates are greater than or equal to the split coordinate. The split axis (in the case of two dimensions, the axes $x$ and $y$) of the coordinate stored alternates for each level of the tree -- if the first level nodes store x-coordinates, the second level nodes store y-coordinates and so on. Each leaf node also represents a region of the space, but it does not store any split coordinate. Instead, it stores a list of the avatars present in that region. Finally, each leaf node is associated to a server of the game. When a server is overloaded, it triggers the load balancing, which uses the kd-tree to readjust the split coordinates that define its region, reducing the amount of content managed by it.

Each node of the tree also stores two other values: capacity and load of the subtree. The load of a non-leaf node is equal to the sum of the load of its children. Similarly, the capacity of a non-leaf node is equal to the sum of the capacity of its children nodes. For the leaf nodes, these values are the same of the server associated to each one of them. The tree root stores, therefore, the total weight of the game and the total capacity of the server system.

@@

In Algorithm \ref{alg:buildtree}, the $id$ value is used to calculate whether each node has children or not and, in the leaf nodes, it determines the server associated to the region represented by each leaf of the tree. The purpose of this is to create a balanced tree, where the number of leaf nodes on each of the two sub-trees of any node differs, in the maximum, by one. In Figure \ref{fig:kdtree} (a), we have a full kd-tree formed with this simple algorithm and, in Figure \ref{fig:kdtree} (b), an incomplete kd-tree with six-leaf nodes. As we can see, each node of the tree in (b) has two sub-trees whose number of leaf nodes differs by one in the worst case.

@@

However, this distribution is not optimal, for the load imposed by the players depends on how they are interacting with one another. For example, if the avatars of two players are distant from each other, there will be probably no interaction between them and, therefore, the server will need only to update each one of them about the outcome of his own actions -- for these, the growth in the number of messages is linear to the number of players. On the other hand, if the avatars are close to each other, each player should be updated not only about the outcome of his own actions but also about the actions of every other player -- in this case, the number of messages may grow quadratically to the number of players \mbox{(Figure \ref{fig:load})}. For this reason, it is not sufficient only to consider the number of players to divide them among the servers.

@@

For this, two nested loops are used to sweep the avatars array, where each of the avatars contains a $load$ variable initialized with zero. As the vector is sorted, the inner loop may start from an index before which it is known that no avatar $a_j$ has relevance to that being referenced in the outer loop, $a_i$. It is used a variable $begin$, with initial value of zero: if the coordinate of $a_j$ is smaller than that of $a_i$, with a difference greater than the maximum view range of the avatars, the variable $begin$ is incremented. For each $a_j$ which is at a distance smaller than the maximum view range, the $load$ of $a_i$ is increased according to the relevance of $a_j$ to $a_i$. When the inner loop reaches an avatar $a_j$, such that its coordinate is greater than that of $a_i$, with a difference greater than the view range, the outer loop moves immediately to the next step, incrementing $a_i$ and setting the value of $a_j$ to that stored in $begin$ (Figure \ref{fig:sweep}).

@@

Each server maintains an array of the avatars located in the region managed by it, sorted according to the $x$ coordinate. Also, each element of the array stores a pointer to another element, forming a chained list that is ordered according to the $y$ coordinated of the avatars (Figure \ref{fig:vectorxlisty}). By maintaining a local sorted avatar list on each server, the time required for balancing the load is somewhat reduced, for there will be no need for the server performing the rebalance to sort again the avatar lists sent by other servers. It will need only to merge all the avatars lists received from the other servers in an unique list, used to define the limits of the regions, what is done by changing the split coordinates which define the space partitions.

@@

To evaluate the proposed dynamic load balancing algorithm, a virtual environment across which many avatars moved was simulated. Starting from a random point in the environment, each avatar moved according to the random waypoint model \cite{bettstetter2004spr}. To force a load imbalance and stress the algorithms tested, we defined some \emph{hotspots} -- points of interest to which the avatars moved with a higher probability than to other parts of the map. This way, a higher concentration of avatars was formed in some areas. Although the movement model used is not very realistic in terms of the way the players move their avatars in real games, it was only used to put to test the load balance algorithms simulated. For each algorithm tested, we simulated two situations: one with the presence of hotspots and one without hotspots.

>>>>

%% Author and Affiliation (multiple authors). Use: and between authors
%\author{Manuscript number 60381}
\author{Carlos Eduardo B. Bezerra, João L. D. Comba, Cláudio F. R. Geyer\\
				Instituto de Informática\\
				Universidade Federal do Rio Grande do Sul\\
				Av. Bento Gonçalves, 9500, Porto Alegre
}

%\contactinfo{60381@jems}
\contactinfo{\{carlos.bezera, comba, geyer\}@inf.ufrgs.br}

%% Keywords that describe your work.
\keywords{MMOGs, load balancing, distributed server, kd-trees.}

@@

In order to reduce this cost, several decentralized solutions have been proposed. Some of them use peer-to-peer networks, such as \cite{schiele2007rpp,rieche2007ppb,hampel2006ppa,elrhalibi2005abm,iimura2004zfg,knutsson2004pps}. Others propose the use of a distributed server composed of low-cost nodes connected through the Internet, as in \cite{ng2002msa,chertov:olb,lee2003sdl,assiotis2006dam}. Anyway, in all these approaches, the ``world'', or virtual environment of the game is divided into regions and for every region is assigned a server -- or a group of peers to manage it, when using peer-to-peer networks. Each of these regions must have a content such that the load imposed on the corresponding server is not greater than its capacity.

@@

\cite{ahmed2008mol}, for example, propose a cell-oriented load balancing model. To balance the load, their algorithm finds, first, all clusters of cells that are managed by the overloaded server. The smallest cluster is selected and, from this cluster, it is chosen the cell which has the least interaction with other cells of the same server -- the interaction between two cells A and B is defined by the authors as the number of pairs of avatars interacting with each other, one of them in A and the other one in B. The selected cell is then transferred to the least loaded server, considering ``load'' as the bandwidth used to send state updates to the players whose avatars are positioned in the cells managed by that server. This process is repeated until the server is no longer overloaded or there is no more servers capable of absorbing more load -- in this case, one option could be to reduce the frequency at which state update messages are sent to the players, as suggested by \cite{bezerra2008a3}.

@@

In \cite{bezerra2009lbs}, it is also proposed the division into cells. To perform the division, the environment is represented by a graph (Figure \ref{fig:graph}), where each vertex represents a cell. Every edge in the graph connects two vertices representing neighboring cells. The weight of a vertex is the server's bandwidth occupied to send state updates to the players whose avatars are in the cell represented by that vertex. The interaction between any two cells define the weight of the edge connecting the corresponding vertices. To form the regions, the graph is partitioned using a greedy algorithm: starting from the heaviest vertex, at each step it is added the vertex connected by the heaviest edge to any of the vertices already selected, until the total weight of the partition of the graph -- defined as the sum of the vertices' weights -- reaches a certain threshold related to the total capacity of the server that will receive the region represented by that partition of the graph.

@@

This kind of data structure is generally used in computer graphics. However, as in MMOGs there is geometric information -- such as the position of the avatars in the environment --, space partitioning trees can be used. Moreover, we cand find in the literature techniques for keeping the partitions defined by the tree with a similar ``load''. In \cite{luque2005bpc}, for example, it is sought to reduce the time needed to calculate the collisions between pairs of objects moving through space. The authors propose the use of a BSP (binary space partitioning) tree to distribute the objects in the scene (Figure \ref{fig:bsp}). Obviously, if each object of a pair is completely inserted in a different partition, they do not collide and there is no need to perform a more complex test for this pair. Assuming an initial division, it is proposed by the authors a dynamic readjustment of the tree as objects move, balancing their distribution on the leaf-nodes of the tree and, therefore, minimizing the time required to perform the collision detection. Some of the ideas proposed by the authors may be used in the context of load balancing between servers in MMOGs.

@@

The load balancing approach proposed here is based on two criteria: first, the system should be considered heterogeneous (i.e. every server may have a different amount of resources) and, second, the load on each server is \emph{not} proportional to the number of players connected to it, but to the amount of bandwidth required to send state update messages to them.

This choice is due to the fact that every player sends commands to the server at a constant rate, so the number of messages received by the server per unit time grows linearly with the number of players, whereas the number of state update messages sent by the server may be quadratic, in the worst case.

@@

Each node of the tree represents a region of the space and, moreover, in this node it is stored a split coordinate. Each one of the two children of that node represents a subdivision of the region represented by the parent node, and one of them represents the sub-region before the split coordinate and the other one, the sub-region containing points whose coordinates are greater than or equal to the split coordinate. The split axis (in the case of two dimensions, the axes $x$ and $y$) of the coordinate stored alternates for every level of the tree -- if the first level nodes store x-coordinates, the second level nodes store y-coordinates and so on. Every leaf node also represents a region of the space, but it does not store any split coordinate. Instead, it stores a list of the avatars present in that region. Finally, each leaf node is associated to a server of the game. When a server is overloaded, it triggers the load balancing, which uses the kd-tree to readjust the split coordinates that define its region, reducing the amount of content managed by it.

Every node of the tree also stores two other values: capacity and load of the subtree. The load of a non-leaf node is equal to the sum of the load of its children. Similarly, the capacity of a non-leaf node is equal to the sum of the capacity of its children nodes. For the leaf nodes, these values are the same of the server associated to each one of them. The tree root stores, therefore, the total weight of the game and the total capacity of the server system.

@@

In Algorithm \ref{alg:buildtree}, the $id$ value is used to calculate whether each node has children or not and, in the leaf nodes, it determines the server associated to the region represented by each leaf of the tree. The purpose of this is to create a balanced tree, where the number of leaf nodes on each of the two sub-trees of any node differs, in the maximum, by one. In Figure \ref{fig:kdtree} (a), we have a full kd-tree formed with this simple algorithm and, in Figure \ref{fig:kdtree} (b), an incomplete kd-tree with six-leaf nodes. As we can see, every node of the tree in (b) has two sub-trees whose number of leaf nodes differs by one in the worst case.

@@

However, this distribution is not optimal, for the load imposed by the players depends on how they are interacting with one another. For example, if the avatars of two players are distant from each other, there will be probably no interaction between them and, therefore, the server will need only to update every one of them about the outcome of his own actions -- for these, the growth in the number of messages is linear with the number of players. On the other hand, if the avatars are close to each other, each player should be updated not only about the outcome of his own actions but also about the actions of every other player -- in this case, the number of messages may grow quadratically with the number of players \mbox{(Figure \ref{fig:load})}. For this reason, it is not sufficient only to consider the number of players to divide them among the servers.

@@

For this, two nested loops are used to sweep the avatars array, where each of the avatars contains a $load$ variable initialized with zero. As the vector is sorted, the inner loop may start from an index before which it is known that no avatar $a_j$ has relevance to that being referenced in the outer loop, $a_i$. It is used a variable $begin$, with initial value of zero: if the coordinate of $a_j$ is smaller than that of $a_i$, with a difference greater than the maximum view range of the avatars, the variable $begin$ is incremented. For every $a_j$ which is at a distance smaller than the maximum view range, the $load$ of $a_i$ is increased according to the relevance of $a_j$ to $a_i$. When the inner loop reaches an avatar $a_j$, such that its coordinate is greater than that of $a_i$, with a difference greater than the view range, the outer loop moves immediately to the next step, incrementing $a_i$ and setting the value of $a_j$ to that stored in $begin$ (Figure \ref{fig:sweep}).

@@

Every server maintains an array of the avatars located in the region managed by it, sorted according to the $x$ coordinate. Also, each element of the array stores a pointer to another element, forming a chained list that is ordered according to the $y$ coordinated of the avatars (Figure \ref{fig:vectorxlisty}). By maintaining a local sorted avatar list on each server, the time required for balancing the load is somewhat reduced, for there will be no need for the server performing the rebalance to sort again the avatar lists sent by other servers. It will need only to merge all the avatars lists received from the other servers in an unique list, used to define the limits of the regions, what is done by changing the split coordinates which define the space partitions.

@@

To evaluate the proposed dynamic load balancing algorithm, a virtual environment across which many avatars moved was simulated. Starting from a random point in the environment, each avatar moved according to the random waypoint model \cite{bettstetter2004spr}. To force a load imbalance and stress the algorithms tested, we defined some \emph{hotspots} -- points of interest to which the avatars moved with a higher probability than to other parts of the map. This way, a higher concentration of avatars was formed in some areas. Although the movement model used is not very realistic in terms of the way the players move their avatars in real games, it was only used to verify the load balance algorithms simulated. For each algorithm tested, we simulated two situations: one with the presence of hotspots and one without hotspots.
